Software Development in Startup Companies: The Green-field Startup Model

### 2013-02-22 article/main.tex
Formatted to 80 chars line-lenght

% The publisher's ID mark at the bottom of the page is less important with
% Computer Society journal papers as those publications place the marks
% outside of the main text columns and, therefore, unlike regular IEEE
% journals, the available text space is not reduced by their presence.
% If you want to put a publisher's ID mark on the page you can do it like
% this:
%\IEEEpubid{0000--0000/00\$00.00~\copyright~2007 IEEE}
% or like this to get the Computer Society new two part style.
%\IEEEpubid{\makebox[\columnwidth]{\hfill 0000--0000/00/\$00.00~\copyright~2007 IEEE}%
%\hspace{\columnsep}\makebox[\columnwidth]{Published by the IEEE Computer Society\hfill}}
% Remember, if you use this you must call \IEEEpubidadjcol in the second
% column for its text to clear the IEEEpubid mark (Computer Society jorunal
% papers don't need this extra clearance.)

@@

%--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------%
\section{Introduction}
\label{intro}
% Computer Society journal papers do something a tad strange with the very
% first section heading (almost always called "Introduction"). They place it
% ABOVE the main text! IEEEtran.cls currently does not do this for you.
% However, You can achieve this effect by making LaTeX jump through some
% hoops via something like:
%
%\ifCLASSOPTIONcompsoc
%  \noindent\raisebox{2\baselineskip}[0pt][0pt]%
%  {\parbox{\columnwidth}{\section{Introduction}\label{sec:introduction}%
%  \global\everypar=\everypar}}%
%  \vspace{-1\baselineskip}\vspace{-\parskip}\par
%\else
%  \section{Introduction}\label{sec:introduction}\par
%\fi
%
% Admittedly, this is a hack and may well be fragile, but seems to do the
% trick for me. Note the need to keep any \label that may be used right
% after \section in the above as the hack puts \section within a raised box.

@@

Practitioners can use the GSM model to apply presented strategies to speed-up the development by taking in consideration the amount of technical debt accumulated during the first early-stage period. In this regard, we identified several commonalities between the issues related to software development in startups and the research focused at studying technical debt \cite{Nugroho2011,Izurieta2012}. On the other hand, researchers can take advantage from the GSM model to understand how the technical debt influences the future growth of startup companies.

The remainder of this paper is structured as follows:  we provide relevant information about startups in section \ref{backg}. Section \ref{resmet} introduces the research questions and discusses the design and execution of the research process. Results are discussed in section \ref{res:gsm}, which presents the GSM model. Section \ref{res:val}  compares results of this research to literature and frameworks presented in this field. Section \ref{valt} discusses the validity threats considered during this study. Section \ref{sect:theory:impl} discusses the most relevant implications of the GSM model. Finally, this paper concludes in section \ref{conc}.

%MUN: This is maybe only a minor concern, but it may be relevant for certain reviewers: The references you use
% in the introduction are not very recent. Look at the list of papers [1-15], most of them are published before 2005.
% I understand that you use some references to strengthen your general arguments (which can refer to older publications),
% but your introduction should also show the relevance of your research right now. This is best shown by referring to
% recent (max 5 years old) literature. Think about if you can add more recent (2010-now), well published, papers in the
% introduction.

@@

%\hfill January 11, 2007
%--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------%
\section{Background}
\label{backg}

@@

In this context, software engineering (SE) faces complex and multifaceted obstacles in understanding how to manage development processes. Sutton defines startups as creative and flexible in nature and reluctant to introduce process or bureaucratic measures which may result ineffective \cite{Sutton2000}. Also Bach refers to startups as \textit{``a bunch of energetic and committed people without defined development processes''} \cite{Bach1998}. In fact, startups have very limited resources and typically wish to use them to support product development instead of establishing processes \cite{Coleman2008, Heitlager2007}. Some attempts to taylor lightweight processes to startups reported basic failure of their application: \textit{``Everyone is busy, and software engineering practices are often one of the first places developers cut corners''} \cite{Martin2007}. Rejecting the notion of repeatable and controlled processes, startups prominently take advantage of unpredictable, reactive and low-precision \cite{
surviving-os-cockburn} engineering practices \cite{Sutton2000, Tanabian2005, Chorev2006, Kakati2003}.

@@

%--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------%
\subsection{General lack of research in startups}
Despite some studies tried to address the above mentioned issues, a SMS \cite{SMS} revealed only a few SE works in this specific area, as confirmed by other studies \cite{Coleman2008, Coleman2008a, Coleman2007, Sutton2000}. Moreover, the studies, identified in \cite{SMS}, appear to be highly fragmented and spread across different areas rather than constituting a consistent \textit{body of knowledge}.

@@

%--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------%
\subsection{Software development in startups}

The word \textit{startup} appeared in the SE literature for the first time in 1994 in an article written by Carmel \cite{Camel1994a} where he studied the \textit{time-to-completion in a young package firm}. %MUN: If this is a quote from
%Carmel, put it in quotation marks, like you did in the previous section with Suttons quote. Generally, check that you
%are consistent how you use quotations throughout the paper.
He noticed how these companies were particularly innovative and successful, advocating for the need of more research investigating software development practices so as to replicate success and try to transfer it to other technology sectors.

@@

%--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------%
\subsection{Software process improvement in startups}

@@

Additionally, the role of SPI has been neglected because seen as an obstacle to the development team's creativity and flexibility \cite{Coleman2008a}, and to the need of a quick delivering product process environment. %MUN: the second
%part of the sentence is just a statement without reference. If it refers to a statement by Coleman, put his reference
%at the end OR find the correct reference OR delete the statement.
Product quality is left aside in favor of minimal and suitable functionalities to shorten the time-to-market. As reported in two studies of Mater and Mirel \cite{Mater2000, Mirel2000}, quality aspects, mostly taken in consideration in internet startups, are oriented to usability and scalability, even though the market and application type heavily influences the quality-demand \cite{Coleman2008, Kim2005}.

To maintain the development activities, oriented to limited but suitable functionality, studies suggest to externalize the complexity of parts of the project to third party solutions by means of outsourcing activities, software reuse and open-source strategies \cite{Hanna2010, Jansen2008, Wall2001, Bean2005}. %MUN: Can you split up the references? I.e. write:
%"outsourcing activities \cite{Bla}, software reuse \cite{Bla}, and open-source strategies \cite{Bla}.

%--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------%
\subsection{Technical debt}
Notwithstanding, a new area of the SE research, trying to tackle the problem of the \textit{technical debt},  brings and encompasses different implications in studying development in software startups \cite{Tom2013}. The metaphoric neologism was originally introduced by Cunningham in 1992 \cite{TechnicalDebtCunn} and has recently attracted the attention of SE researchers\footnote{To attest the fact that \textit{technical debt} is gaining traction among researchers, we mention two important contributions which characterize the \textit{``debt landscape''} : \cite{Nugroho2011,Izurieta2012} and a dedicated workshop \cite{workshopDebt} organized by the Software Engineering Institute  and ICSE.}. The concept of \textit{technical debt} is well illustrated in \cite{Brown:2010:MTD:1882362.1882373} : \textit{``The idea is that developers sometimes accept compromises in a system in one dimension (e.g., modularity) to meet an urgent demand in some other dimension (e.g., a deadline), and that such compromises incur a 
``debt'
' on which ``interest'' has to be paid and which the ``principal'' should be repaid at some point for the long-term health of the project''}.  The compromise between high-speed and high-quality engineering is faced daily by software startups, not only in terms of architecture design but in multifaceted aspects (weak project management, testing, process control).

@@

\begin{compactitem}
\item Software development strategy: the overall approach adopted by the company to carry out the product development.
\item Engineering activities: the activities needed to bring a product from idea to market. Traditional engineering activities are, among others, requirement engineering, design, architecture, implementation, testing.
\item Engineering elements: any practice, tool and artifacts contributing and supporting the engineering activities.
\item Quality attributes: those overall factors that affect run-time behavior, system design, and user experience. They represent areas of concern that have the potential for applications to impact across layers and tiers. Some of these attributes are related to the overall system design, while others are specific to run time, design time, or user centric issues \cite{Microsoft2009}.
\item Growth: an increase in the company size respect to the initial conditions in terms of either employees or users/customers, and product complexity in terms of handling an increasing number of feature requests.
\item Software product: any software product and software service.
\end{compactitem}
%--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------%
\section{Research methodology}
\label{resmet}

\begin{figure*}[!t]
\centering
\includegraphics[width=4in]{figures/completemethodology2}
\caption{Grounded theory study process overview}\label{fig:gt:completemethodology}
\end{figure*}
%MUN: I think you can increase the font size in the figure a bit. Try thereby to keep the 
%height of the figure, expanding however the width (use horizontal space since it is a two 
%column figure). Change also in figure:
% - Write "Data Analysis" horizontally
% - Change names of macro-phases to upper-case, i.e. Design and Execution, Data Collection, Data Analysis, etc.
% - Add section number to macro-phase name, i.e. 3.1 Design and Execution, 3.2 Data Collection
% - Change "Discussion and Validation" into "Theory Validation" as you name it in the text.
% Also, add a section (3.5) that describes shortly what you did in "Theory Validation"

@@

\textit{Grounded Theory} methodology is described by Robson as \textit{``the best approach to answer the question - What is going on here?''} and defined by its creators \cite{ColinRobson2009} as \textit{``a set of well-developed categories ( e.g. themes, concepts ) that are systematically interrelated through statements of relations to form a theoretical framework that explains some relevant social, psychological, educational, nursing or other phenomenon''} .

@@

%--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------%
\subsection{Design and Execution}
\label{desex}

@@

All companies, except C10, were founded within the last three years (2009-2012), %MUN: I've shortened the text and
%added absolute numbers. Is 2009-2012 correct?
by an average of 3 starting members, which were in majority developers. Moreover, the number of current employees shows how, to different degrees, companies expanded the initial teams. All companies, except C5, brought their first product to market in less than 6 months from the idea conception. The products consist of web-applications, launched in different nations and market sectors. The growing team size and the publicly available data suggest an healthy status of the businesses.

@@

The follow-up questionnaires were designed to capture additional data, gather missing information and confirm interview results by triangulation. The questionnaires have been tailored to each startup, partially taking advantage of the repertory grid principles \cite{Edwards2009}. To obtain better chances of quickly obtaining responses, questionnaires have been sent to respondents immediately after transcribing the interview results. Two weeks after the conclusion of the last interview, we closed our follow-up questionnaires and collected the data. %MUN: I don't understand this: you
% collected the data from questionnaires after ALL interviews have been finished, but on the other hand you 
% said the overall process was iterative (data analysis and triangulation let themes emerge which in turn were used
% in subsequent interviews).

%--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------%
\subsection{Data collection}
The data collection was designed to allow triangulation, which integrates multiple data sources converging on the same phenomenon. After transcribing the interview, we extracted conceptual relations, integrating them with the questionnaire's results. A well structured case-study database allowed us to easily retrieve and seek for information, assembling the evidence from the case study reports, as described also in \cite{Yin1994}. The database has been stored and constructed using the qualitative data analysis software package AtlasTI\footnote{Available online at \url{http://www.atlasti.com/} .} - one of the most suitable tools for GT \cite{Coleman2007}. 
We overlapped interviews with questionnaire results to adjust the data collection and take advantage of emergent themes and reveal possible inconsistencies. The data was analyzed simultaneously and with flexibility in mind in such a way that adjustments were made according to the emerging findings.
%--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------%
\subsection{Data analysis}
Before starting the analysis, a data ordering procedure was necessary since interviews were spread across a multitude of topics. Therefore transcripts have been structured in thematic areas accordingly to different topic cards used during the interviews. We proceeded horizontally between same thematic areas of transcripts to be able to identify a better number of similar concepts, rather than going through an entire transcript at one time.

@@

%--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------%
\subsection{Theory generation}

@@

\subsection{Theory Validation}
%MUN: put here what I wrote in the comment in Section "Theory validation" (i.e. the section following the GSM)

%--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------%

@@

By analyzing the interview transcripts with the GT process we extracted the information necessary to form the theoretical model. The initial coding process (open and in-vivo) provided 630 unique codes \footnote{Available at \url{https://github.com/adv0r/BTH-Interview-Package}.}, grounded to 1295 citations in the transcripts, divided according to the thematic areas. The thematic areas were divided according to the development phases designed for the interview process. The raw codes constitute the empirical foundation of the theoretical categories, which has been used to form the Green-field startup model (GSM).

The GSM captures the underlying phenomenon of software development in early-stage startups. The framework is formed by 128 sub-categories, clustered in 35 groups, and finally in 7 categories at the highest level of abstraction. By the means of the GSM we provide explanations of the development strategies and engineering activities undertaken by startups.
%--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------%
\subsection{Framework overview}
\label{res:gsm:frmov}
The main concepts representing the underlying phenomenon have been grouped together to form high-level categories. Figure \ref{fig:gsm} shows the network of causal relationships (represented by arrows) between categories (represented by blocks). Each node is linked to subsequent nodes which are called successors, and preceding nodes, called predecessors. The relations between them are denoted by the direction of the interconnecting arrows. The network is read from left to right.

@@

In forthcoming the explanation of the GSM we emphasize labels of theoretical categories using the \textit{italic} font style. For the main categories shown in Figure \ref{fig:gsm}, when necessary we also make use of identifiers (i.e. CATx) to facilitate the reading of this section.
The network is centered around the core category, \textit{speed-up development}, which is the most interconnected node in the framework reflecting the fact that \textit{``it is the one (category) with the greatest explanatory power''}  \cite{Strauss1998}. The meaning of the labels on the three arrows that reach the core category (efficiency, effectiveness and performance) is discussed in detail in section \ref{valt}. %MUN: this refers to the validity threats section. Do you mean rather the "Theory Validation" section?

@@

The \textit{severe lack of resources} forces the company to focus on implementing an essential set of functionalities and it is one of the main reasons why the \textit{product quality has low priority} respect to other more urgent needs\footnote{There are some exceptions where the quality aspects actually matter and such cases will be discussed in section \ref{valt} %MUN: this refers to the validity threats section. Do you mean rather the "Theory Validation" section?
.}. At the same time, to be able to deal with such constraints, startups must depend on a small group of capable and motivated individuals.

@@

In the following subsections we are going to explain in detail all the categories presented in Figure \ref{fig:gsm}, and conclude in subsection \ref{res:gsm:th} with the final theory. In the explanations we use identifiers of the companies presented in Table \ref{t_interviews-stats} (i.e. Cx) to highlight statements recorded from the interviewees.

%--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------%
\subsection{Severe lack of resources}
\label{res:gsm:cat7}
The concept of \textit{severe lack of resources} characterizes the uncertainty of the development strategies in startups and it is composed by three subcategories: \textit{time-shortage}, \textit{limited human resources} and \textit{limited access to expertise}.

@@

The extent to which startups have access to specialized knowledge - both internal and external to the company - is extremely reduced when compared to traditional established software companies. Therefore, in order to partially mitigate the \textit{limited access to expertise}, startups rely on the external aid of mentors or advisors. Under these strict limitations, most of the decisions related to software development are fundamentally trade-off situations (confirming Himola's results \cite{Hilmola2003}).
%--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------%
\subsection{Team is the catalyst of development}
\label{res:gsm:cat4}
Among the different aspects fostering the speed of the development process, the startups' main focus is on the characteristics of the initial team.

@@

A \textit{limited access to expertise} forces the team to rely mainly on their personal abilities, even though asking opinions to mentors is reported to be a viable practice to maintain feasible objectives. Furthermore \textit{teams work under constant pressure}, \textit{``up to 16 hours per day''}, %MUN: add Cx to this statement since it is a citation?
mainly constrained by a tight \textit{time shortage}.

Finally, startups present founders-centric structures, and especially in the early-stage, it is clear that \textit{CTO/CEO background has high-impact} on the company's development approach. For instance, in case of an academic background, the CTO encourages the introduction of some architectural design before the development phase. Moreover, even though the CTO/CEO starts guiding initially the development process, most of the decision are taken consensually by all members of the team. Then, the CTO/CEO decides only in particular situations where some conflicts occur.
%--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------%
\subsection{Evolutionary approach}
\label{res:gsm:cat2}
Among different approaches to development, startups prefer to build an initial prototype and iteratively refine it over time, similarly to the concept of \textit{``evolutionary prototyping''} \cite{EvProt}.  The reason of using this approach is that startups want to validate the product against the market as soon as possible, finding the proper product/market fit. Indeed, they can focus on developing only parts of the system they want to validate instead of working on developing a whole new system. Then, as the prototype is released, users detect opportunities for new functionalities and improvements, providing their feedback to developers.

@@

%--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------%
\subsection{Product quality has low priority}
\label{res:gsm:cat3}
The interests of software startups, related to the product, are mainly concentrated on building a  \textit{limited number of suitable functionalities} rather than fulfilling restrictive non-functional requirements. This strategy allows them to quickly release simple products with less need of preliminary architectural studies.

@@

%--------------------------%%--------------------------%%--------------------------%%--------------------------%%--------------------------%%--------------------------%%--------------------------%
\section{Theory Validation}
\label{res:val}

>>>>

% The publisher's ID mark at the bottom of the page is less important with
% Computer Society journal papers as those publications place the marks
% outside of the main text columns and, therefore, unlike regular IEEE
% journals, the available text space is not reduced by their presence.
% If you want to put a publisher's ID mark on the page you can do it like
% this:
%\IEEEpubid{0000--0000/00\$00.00~\copyright~2007 IEEE}
% or like this to get the Computer Society new two part style.
%\IEEEpubid{\makebox[\columnwidth]{\hfill 0000--0000/00/\$00.00~\copyright~2007  
% IEEE}%
%\hspace{\columnsep}\makebox[\columnwidth]{Published by the IEEE Computer  
% Society\hfill}}
% Remember, if you use this you must call \IEEEpubidadjcol in the second
% column for its text to clear the IEEEpubid mark (Computer Society jorunal
% papers don't need this extra clearance.)

@@

%------------------------------------------------------------------------------%
\section{Introduction}
\label{intro}
% Computer Society journal papers do something a tad strange with the very
% first section heading (almost always called "Introduction"). They place it
% ABOVE the main text! IEEEtran.cls currently does not do this for you.
% However, You can achieve this effect by making LaTeX jump through some
% hoops via something like:
%
%\ifCLASSOPTIONcompsoc
%  \noindent\raisebox{2\baselineskip}[0pt][0pt]%
%  {\parbox{\columnwidth}{\section{Introduction}\label{sec:introduction}%
%  \global\everypar=\everypar}}%
%  \vspace{-1\baselineskip}\vspace{-\parskip}\par
%\else
%  \section{Introduction}\label{sec:introduction}\par
%\fi
%
% Admittedly, this is a hack and may well be fragile, but seems to do the
% trick for me. Note the need to keep any \label that may be used right
% after \section in the above as the hack puts \section within a raised box.

@@

Practitioners can use the GSM to apply presented strategies to speed-up the 
development by taking in consideration the amount of technical debt accumulated 
during the first early-stage period. In this regard, we identified several 
commonalities between the issues related to software development in startups and 
the research focused at studying technical debt \cite{Nugroho2011,Izurieta2012}. 
On the other hand, researchers can take advantage from the GSM to 
understand how the technical debt influences the future growth of startup 
companies.

The remainder of this paper is structured as follows:  we provide relevant 
information about startups in section \ref{backg}. Section \ref{resmet} 
introduces the research questions and discusses the design and execution of the 
research process. Results are discussed in section \ref{res:gsm}, which presents 
the GSM. Section \ref{res:val}  compares results of this research to 
literature and frameworks presented in this field. Section \ref{valt} discusses 
the validity threats considered during this study. Section 
\ref{sect:theory:impl} discusses the most relevant implications of the GSM. 
Finally, this paper concludes in section \ref{conc}.

%MUN: This is maybe only a minor concern, but it may be relevant for certain 
%reviewers: The references you use in the introduction are not very recent. 
%Look at the list of papers [1-15], most of them are published before 2005.
%I understand that you use some references to strengthen your general arguments 
%(which can refer to older publications), but your introduction should also 
%show the relevance of your research right now. This is best shown by 
%referring to recent (max 5 years old) literature. Think about if you can add 
%more recent (2010-now), well published, papers in the introduction.

@@

%\hfill January 11, 2007
%------------------------------------------------------------------------------%
\section{Background}
\label{backg}

@@

In this context, software engineering (SE) faces complex and multifaceted 
obstacles in understanding how to manage development processes. Sutton defines 
startups as creative and flexible in nature and reluctant to introduce process 
or bureaucratic measures which may result ineffective \cite{Sutton2000}. Also 
Bach refers to startups as \textit{``a bunch of energetic and committed people 
without defined development processes''} \cite{Bach1998}. In fact, startups have 
very limited resources and typically wish to use them to support product 
development instead of establishing processes \cite{Coleman2008, Heitlager2007}. 
Some attempts to taylor lightweight processes to startups reported basic failure 
of their application: \textit{``Everyone is busy, and software engineering 
practices are often one of the first places developers cut corners''} 
\cite{Martin2007}. Rejecting the notion of repeatable and controlled processes, 
startups prominently take advantage of unpredictable, reactive and low-precision 
\cite{surviving-os-cockburn} engineering practices \cite{Sutton2000, 
Tanabian2005, Chorev2006, Kakati2003}.

@@

%------------------------------------------------------------------------------%
\subsection{General lack of research in startups}
Despite some studies tried to address the above mentioned issues, a SMS 
\cite{SMS} revealed only a few SE works in this specific area, as confirmed by 
other studies \cite{Coleman2008, Coleman2008a, Coleman2007, Sutton2000}. 
Moreover, the studies, identified in \cite{SMS}, appear to be highly fragmented 
and spread across different areas rather than constituting a consistent 
\textit{body of knowledge}.

@@

%------------------------------------------------------------------------------%
\subsection{Software development in startups}

The word \textit{startup} appeared in the SE literature for the first time in 
1994 in an article written by Carmel \cite{Camel1994a} where he studied the 
\textit{time-to-completion in a young package firm}. %MUN: If this is a quote 
%from Carmel, put it in quotation marks, like you did in the previous section 
%with Suttons quote. Generally, check that you %are consistent how you use 
%quotations throughout the paper.
He noticed how these companies were particularly innovative and successful, 
advocating for the need of more research investigating software development 
practices so as to replicate success and try to transfer it to other technology 
sectors.

@@

%------------------------------------------------------------------------------%
\subsection{Software process improvement in startups}

@@

Additionally, the role of SPI has been neglected because seen as an obstacle to 
the development team's creativity and flexibility \cite{Coleman2008a}, and to 
the need of a quick delivering product process environment. %MUN: the second
%part of the sentence is just a statement without reference. If it refers to a 
%statement by Coleman, put his reference at the end OR find the correct 
%reference OR delete the statement.
Product quality is left aside in favor of minimal and suitable functionalities 
to shorten the time-to-market. As reported in two studies of Mater and Mirel 
\cite{Mater2000, Mirel2000}, quality aspects, mostly taken in consideration in 
internet startups, are oriented to usability and scalability, even though the 
market and application type heavily influences the quality-demand 
\cite{Coleman2008, Kim2005}.

To maintain the development activities, oriented to limited but suitable 
functionality, studies suggest to externalize the complexity of parts of the 
project to third party solutions by means of outsourcing activities, software 
reuse and open-source strategies \cite{Hanna2010, Jansen2008, Wall2001, 
Bean2005}. %MUN: Can you split up the references? I.e. write:
%"outsourcing activities \cite{Bla}, software reuse \cite{Bla}, and open-source 
%strategies \cite{Bla}.

%------------------------------------------------------------------------------%
\subsection{Technical debt}
Notwithstanding, a new area of the SE research, trying to tackle the problem of 
the \textit{technical debt},  brings and encompasses different implications in 
studying development in software startups \cite{Tom2013}. The metaphoric 
neologism was originally introduced by Cunningham in 1992 
\cite{TechnicalDebtCunn} and has recently attracted the attention of SE 
researchers\footnote{To attest the fact that \textit{technical debt} is gaining 
traction among researchers, we mention two important contributions which 
characterize the \textit{``debt landscape''} : \cite{Nugroho2011,Izurieta2012} 
and a dedicated workshop \cite{workshopDebt} organized by the Software 
Engineering Institute  and ICSE.}. The concept of \textit{technical debt} is 
well illustrated in \cite{Brown:2010:MTD:1882362.1882373} : \textit{``The idea 
is that developers sometimes accept compromises in a system in one dimension 
(e.g., modularity) to meet an urgent demand in some other dimension (e.g., a 
deadline), and that such compromises incur a ``debt'' on which ``interest'' has 
to be paid and which the ``principal'' should be repaid at some point for the 
long-term health of the project''}.  The compromise between high-speed and 
high-quality engineering is faced daily by software startups, not only in terms 
of architecture design but in multifaceted aspects (weak project management, 
testing, process control).

@@

\begin{compactitem}
\item Software development strategy: the overall approach adopted by the 
company to carry out the product development.
\item Engineering activities: the activities needed to bring a product from 
idea to market. Traditional engineering activities are, among others, 
requirement engineering, design, architecture, implementation, testing.
\item Engineering elements: any practice, tool and artifacts contributing and 
supporting the engineering activities.
\item Quality attributes: those overall factors that affect run-time behavior, 
system design, and user experience. They represent areas of concern that have 
the potential for applications to impact across layers and tiers. Some of these 
attributes are related to the overall system design, while others are specific 
to run time, design time, or user centric issues \cite{Microsoft2009}.
\item Growth: an increase in the company size respect to the initial conditions 
in terms of either employees or users/customers, and product complexity in terms 
of handling an increasing number of feature requests.
\item Software product: any software product and software service.
\end{compactitem}
%------------------------------------------------------------------------------%
\section{Research methodology}
\label{resmet}

\begin{figure*}[!t]
\centering
\includegraphics[width=4in]{figures/completemethodology2}
\caption{Grounded theory study process overview}
\label{fig:gt:completemethodology}
\end{figure*}
%MUN: I think you can increase the font size in the figure a bit. Try thereby 
%to keep the height of the figure, expanding however the width (use horizontal 
%space since it is a two column figure). Change also in figure:
% - Write "Data Analysis" horizontally
% - Change names of macro-phases to upper-case, i.e. Design and Execution, 
%   Data Collection, Data Analysis, etc.
% - Add section number to macro-phase name, i.e. 3.1 Design and Execution, 3.2 
%   Data Collection
% - Change "Discussion and Validation" into "Theory Validation" as you name it 
%   in the text.
%Also, add a section (3.5) that describes shortly what you did in "Theory 
%Validation"

@@

\textit{Grounded Theory} methodology is described by Robson as \textit{``the 
best approach to answer the question - What is going on here?''} and defined by 
its creators \cite{ColinRobson2009} as \textit{``a set of well-developed 
categories ( e.g. themes, concepts ) that are systematically interrelated 
through statements of relations to form a theoretical framework that explains 
some relevant social, psychological, educational, nursing or other 
phenomenon''}.

@@

%------------------------------------------------------------------------------%
\subsection{Design and Execution}
\label{desex}

@@

All companies, except C10, were founded within the last three years (2009-2012), 
%MUN: I've shortened the text and added absolute numbers. Is 2009-2012 correct?
by an average of 3 starting members, which were in majority developers. 
Moreover, the number of current employees shows how, to different degrees, 
companies expanded the initial teams. All companies, except C5, brought their 
first product to market in less than 6 months from the idea conception. The 
products consist of web-applications, launched in different nations and market 
sectors. The growing team size and the publicly available data suggest an 
healthy status of the businesses.

@@

The follow-up questionnaires were designed to capture additional data, gather 
missing information and confirm interview results by triangulation. The 
questionnaires have been tailored to each startup, partially taking advantage of 
the repertory grid principles \cite{Edwards2009}. To obtain better chances of 
quickly obtaining responses, questionnaires have been sent to respondents 
immediately after transcribing the interview results. Two weeks after the 
conclusion of the last interview, we closed our follow-up questionnaires and 
collected the data. %MUN: I don't understand this: you
%collected the data from questionnaires after ALL interviews have been 
%finished, but on the other hand you said the overall process was iterative 
%(data analysis and triangulation let themes emerge which in turn were used
%in subsequent interviews).

%------------------------------------------------------------------------------%
\subsection{Data collection}
The data collection was designed to allow triangulation, which integrates 
multiple data sources converging on the same phenomenon. After transcribing the 
interview, we extracted conceptual relations, integrating them with the 
questionnaire's results. A well structured case-study database allowed us to 
easily retrieve and seek for information, assembling the evidence from the case 
study reports, as described also in \cite{Yin1994}. The database has been stored 
and constructed using the qualitative data analysis software package 
AtlasTI\footnote{Available online at \url{http://www.atlasti.com/}.} - one of 
the most suitable tools for GT \cite{Coleman2007}. 
We overlapped interviews with questionnaire results to adjust the data 
collection and take advantage of emergent themes and reveal possible 
inconsistencies. The data was analyzed simultaneously and with flexibility in 
mind in such a way that adjustments were made according to the emerging 
findings.
%------------------------------------------------------------------------------%
\subsection{Data analysis}
Before starting the analysis, a data ordering procedure was necessary since 
interviews were spread across a multitude of topics. Therefore transcripts have 
been structured in thematic areas accordingly to different topic cards used 
during the interviews. We proceeded horizontally between same thematic areas of 
transcripts to be able to identify a better number of similar concepts, rather 
than going through an entire transcript at one time.

@@

%------------------------------------------------------------------------------%
\subsection{Theory generation}

@@

\subsection{Theory Validation}
%MUN: put here what I wrote in the comment in Section "Theory validation" (i.e. 
%the section following the GSM)

%------------------------------------------------------------------------------%

@@

By analyzing the interview transcripts with the GT process we extracted the 
information necessary to form the theoretical model. The initial coding process 
(open and in-vivo) provided 630 unique codes\footnote{Available at 
\url{https://github.com/adv0r/BTH-Interview-Package}.}, grounded to 1295 
citations in the transcripts, divided according to the thematic areas. The 
thematic areas were divided according to the development phases designed for the 
interview process. The raw codes constitute the empirical foundation of the 
theoretical categories, which has been used to form the Green-field startup 
model (GSM).

The GSM captures the underlying phenomenon of software development in 
early-stage startups. The framework is formed by 128 sub-categories, clustered 
in 35 groups, and finally in 7 categories at the highest level of abstraction. 
By the means of the GSM we provide explanations of the development strategies 
and engineering activities undertaken by startups.
%------------------------------------------------------------------------------%
\subsection{Framework overview}
\label{res:gsm:frmov}
The main concepts representing the underlying phenomenon have been grouped 
together to form high-level categories. Figure \ref{fig:gsm} shows the network 
of causal relationships (represented by arrows) between categories (represented 
by blocks). Each node is linked to subsequent nodes which are called successors, 
and preceding nodes, called predecessors. The relations between them are denoted 
by the direction of the interconnecting arrows. The network is read from left to 
right.

@@

In forthcoming the explanation of the GSM we emphasize labels of theoretical 
categories using the \textit{italic} font style. For the main categories shown 
in Figure \ref{fig:gsm}, when necessary we also make use of identifiers (i.e. 
CATx) to facilitate the reading of this section.
The network is centered around the core category, \textit{speed-up 
development}, which is the most interconnected node in the framework reflecting 
the fact that \textit{``it is the one (category) with the greatest explanatory 
power''}  \cite{Strauss1998}. The meaning of the labels on the three arrows that 
reach the core category (efficiency, effectiveness and performance) is discussed 
in detail in section \ref{valt}. %MUN: this refers to the validity threats 
%section. Do you mean rather the "Theory Validation" section?

@@

The \textit{severe lack of resources} forces the company to focus on 
implementing an essential set of functionalities and it is one of the main 
reasons why the \textit{product quality has low priority} respect to other more 
urgent needs\footnote{There are some exceptions where the quality aspects 
actually matter and such cases will be discussed in section \ref{valt}%MUN: 
%this refers to the validity threats section. Do you mean rather the "Theory 
%Validation" section?
.}. At the same time, to be able to deal with such constraints, startups must 
depend on a small group of capable and motivated individuals.

@@

In the following subsections we explain in detail all the categories presented 
in Figure \ref{fig:gsm}, and conclude in subsection \ref{res:gsm:th} with the 
final theory. In the explanations we use identifiers of the companies presented 
in Table \ref{t_interviews-stats} (i.e. Cx) to highlight statements recorded 
from the interviewees.

%------------------------------------------------------------------------------%
\subsection{Severe lack of resources}
\label{res:gsm:cat7}
The concept of \textit{severe lack of resources} characterizes the uncertainty 
of the development strategies in startups and it is composed by three 
subcategories: \textit{time-shortage}, \textit{limited human resources} and 
\textit{limited access to expertise}.

@@

The extent to which startups have access to specialized knowledge - both 
internal and external to the company - is extremely reduced when compared to 
traditional established software companies. Therefore, in order to partially 
mitigate the \textit{limited access to expertise}, startups rely on the external 
aid of mentors or advisors. Under these strict limitations, most of the 
decisions related to software development are fundamentally trade-off 
situations (confirming Himola's results \cite{Hilmola2003}).
%------------------------------------------------------------------------------%
\subsection{Team is the catalyst of development}
\label{res:gsm:cat4}
Among the different aspects fostering the speed of the development process, the 
startups' main focus is on the characteristics of the initial team.

@@

A \textit{limited access to expertise} forces the team to rely mainly on their 
personal abilities, even though asking opinions to mentors is reported to be a 
viable practice to maintain feasible objectives. Furthermore \textit{teams work 
under constant pressure}, \textit{``up to 16 hours per day''}, %MUN: add Cx to 
%this statement since it is a citation? 
mainly constrained by a tight \textit{time shortage}.

Finally, startups present founders-centric structures, and especially in the 
early-stage, it is clear that \textit{CTO/CEO background has high-impact} on the 
company's development approach. For instance, in case of an academic background, 
the CTO encourages the introduction of some architectural design before the 
development phase. Moreover, even though the CTO/CEO starts guiding initially 
the development process, most of the decision are taken consensually by all 
members of the team. Then, the CTO/CEO decides only in particular situations 
where some conflicts occur.
%------------------------------------------------------------------------------%
\subsection{Evolutionary approach}
\label{res:gsm:cat2}
Among different approaches to development, startups prefer to build an initial 
prototype and iteratively refine it over time, similarly to the concept of 
\textit{``evolutionary prototyping''} \cite{EvProt}. The reason of using this 
approach is that startups want to validate the product against the market as 
soon as possible, finding the proper product/market fit. Indeed, they can focus 
on developing only parts of the system they want to validate instead of working 
on developing a whole new system. Then, as the prototype is released, users 
detect opportunities for new functionalities and improvements, providing their 
feedback to developers.

@@

%------------------------------------------------------------------------------%
\subsection{Product quality has low priority}
\label{res:gsm:cat3}
The interests of software startups, related to the product, are mainly 
concentrated on building a  \textit{limited number of suitable functionalities} 
rather than fulfilling restrictive non-functional requirements. This strategy 
allows them to quickly release simple products with less need of preliminary 
architectural studies.

@@

%------------------------------------------------------------------------------%
\section{Theory Validation}
\label{res:val}
